---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MultiOutbreaks

<!-- badges: start -->
<!-- badges: end -->

MultiOutbreaks is created to implement a Bayesian copula-based modelling of multi-type spatio-temporal infectious disease outbreak data as presented in [(Adeoye et al., 2026)](https://arxiv.org). The package is built with different functionalities including model simulations, Bayesian inference methods, model comparison methods, plotting of results, etc.

# Dependencies

<!-- badges: start -->
<!-- badges: end -->

MultiOutbreaks features a complementary HMC sampling which depends on Stan through its R interface, cmdstanr. If not already installed, you can do so using

``` r
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
library(cmdstanr)
install_cmdstan()
```

## Installation

You can install the development version of MultiOutbreaks from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("Matthewadeoye/MultiOutbreaks")
```

## Simulation, visualization, inference, and outbreak detection.

An example showing how to simulate and plot multi-type space-time data from model-0 described in [(Adeoye et al., 2026)](https://arxiv.org):

```{r simulation}
library(MultiOutbreaks)
sim_adjmat<- matrix(0, nrow = 9, ncol = 9)
uppertriang<- c(1,0,1,0,0,0,0,0,1,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,0,0,1,0,1,0,0,0,1,1,0,1)
gdata::upperTriangle(sim_adjmat, byrow=TRUE)<- uppertriang
gdata::lowerTriangle(sim_adjmat, byrow=FALSE)<- uppertriang
set.seed(0); mod0<- simulateMultiModel(Modeltype = 0, time = 60, nstrain = 5, adj.matrix = sim_adjmat)
multitypeFig(y=mod0[["y"]])
```

```{r figures, echo = FALSE}
library(MultiOutbreaks)
sim_adjmat<- matrix(0, nrow = 9, ncol = 9)
uppertriang<- c(1,0,1,0,0,0,0,0,1,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,0,0,1,0,1,0,0,0,1,1,0,1)
gdata::upperTriangle(sim_adjmat, byrow=TRUE)<- uppertriang
gdata::lowerTriangle(sim_adjmat, byrow=FALSE)<- uppertriang
set.seed(0); mod0<- simulateMultiModel(Modeltype = 0, time = 60, nstrain = 5, adj.matrix = sim_adjmat)
multitypeFig(y=mod0[["y"]])
```

Inference using either Hamiltonian Monte Carlo or a bespoke MCMC algorithm via MultiOutbreaks:

```{r inference}
#HMCfit0<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 0, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = TRUE, GPU = FALSE)
#mcmc.plot(HMCfit0)
#inf.plot(HMCfit0)

#MCMCfit0<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 0, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = FALSE)
#mcmc.plot(MCMCfit0)
#inf.plot(MCMCfit0)
```


Simulation, inference and multi outbreak detection as described in [(Adeoye et al., 2026)](https://arxiv.org/):

```{r outbreak-detection}
library(MultiOutbreaks)
sim_adjmat<- matrix(0, nrow = 9, ncol = 9)
uppertriang<- c(1,0,1,0,0,0,0,0,1,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,0,0,1,0,1,0,0,0,1,1,0,1)
gdata::upperTriangle(sim_adjmat, byrow=TRUE)<- uppertriang
gdata::lowerTriangle(sim_adjmat, byrow=FALSE)<- uppertriang
set.seed(0); mod1<- simulateMultiModel(Modeltype = 1, time = 60, nstrain = 5, adj.matrix = sim_adjmat, B = c(1.2,2.4,1.1,0.4,0.7))
#multitypeFig(y=mod1[["y"]])
#HMCfit1<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 1, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = TRUE, GPU = FALSE)
#mcmc.plot(HMCfit1)
#inf.plot(HMCfit1)

#MarginalProbabilitiesHMC<- Posteriormultstrain.Decoding(y=mod1[["y"]],e_it=mod1[["e_it"]],inf.object = HMCfit1,Modeltype = 1,y_total = NULL,thinningL = 10,burn.in = 2000)
#image(t(MarginalProbabilities[,,1]))

#MCMCfit1<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 1, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = FALSE, GPU = FALSE)
#mcmc.plot(MCMCfit1)
#inf.plot(MCMCfit1)

#MarginalProbabilitiesMCMC<- Posteriormultstrain.Decoding(y=mod1[["y"]],e_it=mod1[["e_it"]],inf.object = MCMCfit1,Modeltype = 1,y_total = NULL,thinningL = 10,burn.in = 2000)
#image(t(MarginalProbabilitiesMCMC[,,1]))
```


Compute model evidence (log marginal likelihood) through Bridge or importance sampling for any model fitted by MultiOutbreaks:

```{r model-evidence}
library(MultiOutbreaks)
sim_adjmat<- matrix(0, nrow = 9, ncol = 9)
uppertriang<- c(1,0,1,0,0,0,0,0,1,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,0,0,1,0,1,0,0,0,1,1,0,1)
gdata::upperTriangle(sim_adjmat, byrow=TRUE)<- uppertriang
gdata::lowerTriangle(sim_adjmat, byrow=FALSE)<- uppertriang
set.seed(0); mod1<- simulateMultiModel(Modeltype = 1, time = 60, nstrain = 5, adj.matrix = sim_adjmat, B = c(1.2,2.4,1.1,0.4,0.7))
#HMCfit1<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 1, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = TRUE, GPU = FALSE)
#ModelEvidenceBridgeSamplingPackage(y=mod1[["y"]],e_it = mod1[["e_it"]],adjmat = sim_adjmat,Modeltype =1 ,inf.object = HMCfit1,y_total = NULL)
#ModelEvidence(y=mod1[["y"]],e_it = mod1[["e_it"]],adjmat = sim_adjmat,Modeltype =1 ,inf.object = HMCfit1, num_samples = 5000, y_total = NULL)

#MCMCfit1<- SMOOTHING_INFERENCE(y=mod0[["y"]], e_it=mod0[["e_it"]], Modeltype = 1, sim_adjmat, MCMC_iterations = 15000, HMC_iterations = 5000, Stan = FALSE, GPU = FALSE)
#ModelEvidenceBridgeSamplingPackage(y=mod1[["y"]],e_it = mod1[["e_it"]],adjmat = sim_adjmat,Modeltype =1 ,inf.object = MCMCfit1,y_total = NULL)
#ModelEvidence(y=mod1[["y"]],e_it = mod1[["e_it"]],adjmat = sim_adjmat,Modeltype =1 ,inf.object = MCMCfit1, num_samples = 5000,y_total = NULL)
```
